---
title: "What Makes a Team Hot : Investigating the Red Sox 13 Game Win Streak"
subtitle: ""
author: "Dashiell Brenner"
date: Jul 16, 2026
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(baseballr)
library(jsonlite)
library(plotly)
```

## Introduction

Baseball is often described as "a game of runs," and perhaps no team in the history of the sport has better encapsulated this saying than the 2026 Boston Red Sox. On July 2nd, with their record reading (37-48) and their playoff odds projected at a bleak 9%, fans across Boston were ready to write off the 2026 season as just another disappointment. And then, seemingly overnight, everything changed. Between July 3rd and July 19th the Red Sox ripped off 13 consecutive wins, a feat which has been accomplished fewer than one-hundred times in MLB history. In the span of about 2-weeks the Red Sox went from a team in the basement of the American League to one of the hottest teams in baseball, with their playoff odds rocketing to above 50%.

A momentum shift this dramatic naturally elicited a variety of questions among baseball analysts. Was this simply a lucky streak, the kind of variance any team can run into over a two-week stretch, or had something substantive changed? Were Red Sox hitters suddenly squaring up the ball with more authority? Had the pitching staff turned a real corner? Which players were most responsible for the dramatic turnaround? Were any specific roster additions responsible for the sudden change? This report sets out to answer these questions with data rather than speculation, by comparing Red Sox hitting and pitching metrics from earlier in the season against the period since the win streak began on July 3rd. Along with exploring whether any roster transactions could help explain the shift in the trajectory of the season. 

## Acquisition

To begin my investigation, I acquired data from two distinct sources: Baseball-Reference and the official MLB Stats API. The first source, Baseball-Reference, provided both the Red Sox's full game-by-game results for the 2026 season and player-level batting and pitching statistics split across the two time windows that this report will compare. Baseball-Reference is operated by Sports Reference LLC, which has compiled trusted and organized baseball statistics since 2000. The second source, MLB's official Stats API, provided records of every roster transaction, trades, call-ups, and injured list moves,involving the Red Sox across the 2026 season. The Stats API is maintained directly by Major League Baseball itself. 

Baseball-Reference data was pulled using the baseballr package's bref_team_results(), bref_daily_batter(), and bref_daily_pitcher() functions; two calls each for batters and pitchers (one for before the streak, one for during it.) The MLB Stats API data was acquired by building the request URL myself (with the Red Sox's team ID and a date range) and parsing the JSON response with the fromJSON() function from the jsonlite package.

```{r}
#| label: acquisition

season_year  <- 2026
season_start <- "2026-03-01"
today_date <- "2026-07-20"
streak_start_date <- as.Date("2026-07-03")

# ---- Source 1a: Baseball-Reference team game log ----

# All Red Sox game outcomes from 2026. One row for date, win/loss, starting pitcher, attendance, etc. 

bos_games <- bref_team_results(Tm = "BOS", year = season_year)

write_csv(bos_games, "data/raw/bos_team_results_2026.csv")

# ---- Source 1b: Baseball-Reference player stats log ----

#In-depth Red Sox player stats divided by before streak vs. after streak.

#Boston batting stats before streak
before_batters <- bref_daily_batter(t1 = season_start, t2 = as.character(streak_start_date - 1)) %>% 
  filter(Team == "Boston")

write_csv(before_batters, "data/raw/bos_batters_before_2026.csv")

#Boston batting stats during streak
streak_batters <- bref_daily_batter(t1 = streak_start_date, t2 = today_date) %>% 
  filter(Team == "Boston")

write_csv(streak_batters, "data/raw/bos_batters_streak_2026.csv")

#Boston pitching stats before streak
before_pitchers <- bref_daily_pitcher(t1 = season_start, t2 = as.character(streak_start_date - 1)) %>% 
  filter(Team == "Boston")

write_csv(before_pitchers, "data/raw/bos_pitchers_before_2026.csv")

#Boston pitching stats after streak
streak_pitchers <- bref_daily_pitcher(t1 = as.character(streak_start_date), t2 = today_date) %>% 
  filter(Team == "Boston")

write_csv(streak_pitchers, "data/raw/bos_pitchers_streak_2026.csv")

# ---- Source 2: Official MLB transaction data using API call ----

#All Red Sox roster changes during 2026

url <- paste0(
  "https://statsapi.mlb.com/api/v1/transactions?teamId=111",
  "&startDate=", season_start,
  "&endDate=", today_date
) #costructs url used for API call using variables defined above

result <- fromJSON(url) #converts raw json text to data frame in r

#removes data frames from within data frame and selects desired variables (included in Aquire rather than Wrangle so that the data frame could be written back to csv in data/raw)
bos_transactions <- result$transactions %>% 
  mutate(
    player_name = person$fullName,
    to_team = toTeam$name,
    from_team = fromTeam$name
  ) %>% 
  select(-person, -toTeam, -fromTeam)

write_csv(bos_transactions, "data/raw/bos_transactions_2026.csv")
```

## Manipulation

A significant amount of tidying and manipulating was required before the data could be analyzed and visualized.  
Tidying and manipulation steps:

- Parsed Baseball-Reference's messy date text (no year, doubleheader markers (1)/(2) ) into a proper Date column
- Converted the win/loss result column into a clean logical variable
- Converted home/away and day/night single-letter codes into labeled factors
- Converted the transactions data's date column into a proper Date type
- Converted nested player and team data frames within the transactions data into plain text columns 
- Grouped transactions by date and summarized them into one count/description per day
- Left-joined the summarized transactions onto the game log by date
- Combined before- and during-streak player stats by binding rows, tagging each half with a period label
- Combined all pitchers together and totaled up their metrics by period
- Reshaped pitching stats so each metric could be compared side by side
- Created filtered data set with only batters with sufficient at-bats in both periods
- Ordered batters by their change in OPS from before to during the streak for graphing 

```{r}
#| label: wrangle

#read back data from data folder
bos_games <- read_csv("data/raw/bos_team_results_2026.csv")
bos_transactions <- read_csv("data/raw/bos_transactions_2026.csv")
bos_batters_before <- read_csv("data/raw/bos_batters_before_2026.csv")
bos_batters_streak <- read_csv("data/raw/bos_batters_streak_2026.csv")
bos_pitchers_before <- read_csv("data/raw/bos_pitchers_before_2026.csv")
bos_pitchers_streak <- read_csv("data/raw/bos_pitchers_streak_2026.csv")

#clean game data
bos_games <- bos_games %>% 
  mutate(
    Date_clean = str_remove(Date, " \\(\\d\\)$"), # strip doubleheader marker (1)/(2)
    game_date = as.Date(paste(Date_clean, Year), format = "%A, %b %d %Y"), #reformat dates 
    win = str_starts(Result, "W"), #win column as factor 
    H_A = factor(H_A, levels = c("H", "A"), labels = c("Home", "Away")), #home/away as factor
    `D/N` = factor(`D/N`, levels = c("D", "N"), labels = c("Day", "Night")) #day/ night as factor
  )

#clean transactions data
bos_transactions <- bos_transactions %>% 
  mutate(date = as.Date(date)) #date as string to date 

#group transactions by date and summarize 
transactions_by_date <- bos_transactions %>% 
  group_by(date) %>% 
  summarize(
    n_transactions = n(),
    transaction_summary = toString(player_name) #vector of player names -> string
    )
  
#join game data and transaction data by date
bos_games <- bos_games %>% 
  left_join(transactions_by_date, by = c("game_date" = "date")) |>
  mutate(n_transactions = replace_na(n_transactions, 0))

#add a variable to distinguish between during vs. before streak
bos_batters_before <- bos_batters_before %>% mutate(period = "before")
bos_batters_streak <- bos_batters_streak %>% mutate(period = "streak")
bos_pitchers_before <- bos_pitchers_before %>% mutate(period = "before")
bos_pitchers_streak <- bos_pitchers_streak %>% mutate(period = "streak")

#binds rows of batting stats before vs. after streak to create comparable data set
bos_batters_compare <- bind_rows(bos_batters_before, bos_batters_streak) %>% 
  mutate(period = factor(period, levels = c("before", "streak"))) #make before vs. during streak a factor

#binds rows of pitching stats before vs. after streak to create comparable data set
bos_pitchers_compare <- bind_rows(bos_pitchers_before, bos_pitchers_streak) %>% 
  mutate(period = factor(period, levels = c("before", "streak")))

#combine all pitchers together and total up their metrics by period
team_pitching <- bos_pitchers_compare %>%
  group_by(period) %>%
  summarize(
    total_IP = sum(IP), # total innings pitched by the whole staff
    total_ER = sum(ER), # earned runs allowed
    total_H  = sum(H),
    total_BB = sum(BB),
    total_SO = sum(SO),
    team_ERA  = 9 * total_ER / total_IP, # team earned run average
    team_WHIP = (total_H + total_BB) / total_IP, # walks + hits per inning
    team_K9   = 9 * total_SO / total_IP, # strikeouts per 9 innings
    team_BB9  = 9 * total_BB / total_IP # walks per 9 innings
  )

#reshape so each stat can be compared side by side
team_pitching_long <- team_pitching %>%
  select(period, team_ERA, team_WHIP, team_K9, team_BB9) %>%
  pivot_longer(cols = starts_with("team_"), names_to = "stat", values_to = "value") %>%
  mutate(stat = recode(stat, team_ERA = "ERA", team_WHIP = "WHIP", team_K9 = "K per 9", team_BB9 = "BB per 9"))

#data set including only players with ABS during and before streak
batters_both <- bos_batters_compare %>% 
  group_by(Name) %>% 
  filter(n_distinct(period) == 2) %>% 
  ungroup() %>% 
  filter(!(Name %in% c("Nate Eaton","Anthony Seigler", "Romy Gonzalez" ))) #not enough ABs during or before streak

#ops changes by batter
ops_change <- batters_both %>% 
  select(Name, period, OPS) %>% 
  pivot_wider(names_from = period, values_from = OPS) %>% 
  mutate(change = streak - before) %>% 
  arrange(change)

#Names arranged by OPS change ascending
ops_order <- ops_change %>% pull(Name)

#Reorder batters_both by ops change for graph
batters_both <- batters_both %>% 
  mutate(Name = factor(Name, levels = ops_order))
```

## Visualization

**Figure 1** compares each Red Sox hitter's OPS (on-base plus slugging, a standard measure of overall offensive production) before the win streak against their OPS during it. Each player is represented by two dots connected by a line, with the darker dot mapping their OPS during the streak and the lighter dot mapping their OPS before the streak. 

```{r}
#| label: visualization

#---- OPS dumbbell chart ----

#dumbell plot comparing pre-streak ops to mid-streak ops for each batter with sufficient ABS
ops_dumbell_plot <- ggplot(batters_both, aes(x = Name, y = OPS)) +
  geom_line(aes(group = Name), color = "grey70", linewidth = 0.8) +
  geom_point(aes(color = period), size = 4) +
  scale_color_manual(values = c(before = "#7FB3D5", streak = "#0C2340"),
                      labels = c("Before streak", "During streak")) +
  labs(
    title = "Red Sox hitters' OPS improved across the board during the win streak",
    subtitle = "Each line connects one player's OPS before vs. during the 13-game winstreak",
    x = NULL, y = "OPS", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

ops_dumbell_plot
```

**Figure 2** is an interactive plot which shows every Red Sox batter's OPS specifically during the win streak, colored by whether they were above or below the league-average OPS of .721. Hover over any point to see which player it represents.

```{r}
#---- Interactive OPS comparison chart Plotly ----
 
league_avg_ops <- 0.721 #league average ops

#Add above vs. below league average variable for graph coloring
bos_batters_streak <- bos_batters_streak %>%
  mutate(ops_group = if_else(OPS > league_avg_ops, "Above league average", "Below league average"))

ops_plot <- ggplot(bos_batters_streak, aes(x = reorder(Name, OPS), y = OPS,
                    color = ops_group,
                    text = paste0(Name, "<br>OPS: ", round(OPS, 3)))) +
  geom_point(size = 4) +
  geom_hline(yintercept = league_avg_ops, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("Above league average" = "#4CAF50", "Below league average" = "#BD3039")) +
  labs(
    title = "Red Sox batters' OPS during the win streak",
    x = NULL, y = "OPS", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

interactive_plot <- ggplotly(ops_plot, tooltip = "text")

interactive_plot
```

**Figure 3** illustrates the improvement of the pitching staff using four standard metrics; ERA, WHIP, strikeouts per nine innings, and walks per nine innings. The graph compares the period before the streak to the period during it using four faceted bar charts. Individual pitcher stats were aggregated into team wide metrics, rather than compared one by one, as most pitchers threw very few innings during the 13 game stretch. 

```{r}
#---- Pitching metric comparison chart ----

# bar chart comparing overall team pitching stats before vs. during the streak
pitchers_bar_chart <- ggplot(team_pitching_long, aes(x = period, y = value, fill = period)) +
  geom_col(width = 0.5) +
  facet_wrap(~ stat, scales = "free_y") + # each stat gets its own scale
  scale_fill_manual(values = c(before = "#7FB3D5", streak = "#0C2340")) +
  labs(
    title = "Team pitching improved across the board during the win streak",
    subtitle = "Aggregate Red Sox pitching stats, before vs. during the 12-game streak",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12)

pitchers_bar_chart
```

## Results

Across nearly every offensive and pitching metric, the Red Sox's dramatic season turnaround is mirrored by substantial statistical improvement. As clearly seen in Figure 1, the offensive performance of the team as a whole skyrocketed during the thirteen-game stretch. Eight of the eleven batters with sufficient at-bats both before and during the streak had a noticeably higher OPS during the streak compared to before it, with the largest improvements coming from Andruw Monasterio (+529) and Tsu Che Cheng (+592). By overall OPS, Wilson Contreres (1.120), Masataka Yoshida (1.117), and Andruw Monasterio (1.162) have been the leading contributors to the Red Sox offensive surge. 

On top of these offensive improvements, the pitching staff has hit their stride as well. As seen in Figure 3, aggregate team pitching stats improved across the board during the win streak: ERA fell from 3.85 to 1.88, WHIP dropped from 1.27 to 0.92, and the staff allowed fewer walks per nine (3.10 vs. 1.96). However, strikeouts per nine did drop marginally from 8.74 to 8.36. Although not included in visualization, the leading contributors in the pitching staff were been Jake Bennett (20.2 IP, 2 ER), Sonny Gray (12.0IP, 2ER), and Payton Tolle (9.2 IP, 1 ER.)

My investigation into the influence of roster transactions did not yield any notable results. No contributing players were called up or traded for near the beginning of the streak. Anthony Seigler, Tsu Che Cheng, and Jake Bennett were all added to the roster in mid-June and have made strong positive influences on team results, but their addition came weeks before the streak actually began on July 3rd. 

## References

The following are sources used beyond class website resources. 

toddrob99. *MLB-StatsAPI Wiki: Endpoints*. GitHub. https://github.com/toddrob99/MLB-StatsAPI/wiki/Endpoints

Petti, B. (2024). *baseballr: Acquiring and analyzing baseball data* [R package]. https://billpetti.github.io/baseballr/

Wharton AI & Analytics Initiative. (2025). How to Create Interactive Plots in R – On-Demand Tutorial. YouTube. https://www.youtube.com/watch?v=q93OfLallDw

GeeksforGeeks. (2024, September 12). How to format mouse over labels using ggplotly in R. GeeksforGeeks. https://www.geeksforgeeks.org/r-language/how-to-format-mouse-over-labels-using-ggplotly-in-r/

Ooms, J. (2014). Getting started with JSON and jsonlite. jsonlite package vignette, CRAN. https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html

Artificial Intelligence was used to help in Initial Scaffolding phase to construct repository and write gitignore. 



